Phylogenetics and the human microbiome.

### 2013-10-01 sbreview.tex
Including figures into the MS.

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{palatino}
\usepackage[hidelinks]{hyperref}
\usepackage[round]{natbib}
\newcommand{\forarxiv}[1]{#1}
\newcommand{\notforarxiv}[1]{}

@@

% code
\newcommand{\program}{\textsf{program}}

@@

\newcommand{\FIGmassTransport}{\
\begin{figure}[ht]
\begin{center}
  \forarxiv{\includegraphics[width=13cm]{mass_transport.pdf}}
\end{center}
\caption{\
  Caption goes here.
}
\label{FIGmassTransport}
\end{figure}
}
\newcommand{\refFIGmassTransport}{1}

@@

The human microbiome is the collection of microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information of microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, in particular the impact of tree-thinking on the methods used in the analysis of human microbiome data, as well as describing current challenges for phylogenetics coming from this type of work.

@@

\textbf{[Figure: \url{http://i.imgur.com/3rs1bND.png}]?}

@@

\textbf{[Figure: \url{http://cl.ly/image/3k3n3W3U443G/ref_tree_bwpd.png} to show a partially abundance weighted PD measure?]}

\subsection{Community comparison using phylogenetics}
The level of similarity between samples or groups of samples is called \emph{beta diversity}.
As with alpha diversity, it is not uncommon to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however phylogenetics-based methods are the most popular.
They are generally variants of the ``UniFrac'' phylogenetic dissimilarity measure \citep[as described and named by][]{LozuponeKnightUniFrac05}.
\citet{kuczynski2010microbial} claim that the UniFrac framework is superior to other methods for community comparison via real data and simulations \citep[for a contrary viewpoint using simulations see][]{schloss2008evaluating}.
% It recovers gradients. \citep{nemergut2011global}

\textbf{[Figure: canonical unifrac explanation picture?]}

@@

\textbf{[Figure: \url{http://cl.ly/image/3K472T0K3S3S/tree_dirtpiles.png}]?}.

@@

\textbf{[Figure: example principal coordinates plot?]}

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect more related work.
Future will continue to bring deeper sequencing on more samples.
The uBiome project\footnote{(\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
Comparative studies will continue to study what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, to Noah Hoffman and Bj\"orn Winckler for providing feedback on the manuscript, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.

\notforarxiv{
\newpage
\section{Figure Legends}
%\FIGmassTransport
\clearpage

>>>>

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{palatino}
\usepackage{wrapfig}
\usepackage[hidelinks]{hyperref}
\usepackage[round]{natbib}
\usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
\newcommand{\forarxiv}[1]{#1}
\newcommand{\notforarxiv}[1]{}

@@

@@

@@

The human microbiome is the collection of microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information of microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.

@@

@@

\subsection{Community comparison using phylogenetics}

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  \capstart
  \vspace{-19pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}

The level of similarity between samples or groups of samples is called \emph{beta diversity}.
As with alpha diversity, it is not uncommon to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however phylogenetics-based methods are the most popular.
They are generally variants of the ``UniFrac'' phylogenetic dissimilarity measure \citep[as described and named by][]{LozuponeKnightUniFrac05}.
\citet{kuczynski2010microbial} claim that the UniFrac framework is superior to other methods for community comparison via real data and simulations \citep[for a contrary viewpoint using simulations see][]{schloss2008evaluating}.
% It recovers gradients. \citep{nemergut2011global}

@@

@@

\newcommand{\dirtpilesLegend}{
    Part of a minimal mass movement to calculate the earth-mover's distance on a phylogenetic tree.
}
\forarxiv{
\begin{wrapfigure}[19]{l}{1.6in}
  \capstart
% \vspace{-24pt}
  \begin{center}
    \includegraphics[width=1.6in]{figures/tree_dirtpiles.pdf}
  \end{center}
  \vspace{-5pt}
  \caption{\dirtpilesLegend}
  \label{fig:dirtpiles}
\end{wrapfigure}
}

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect more related work.
Future will continue to bring deeper sequencing on more samples.
The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
Comparative studies will continue to study what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, to Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.

\notforarxiv{
\newpage
\section{Figure Legends}
%\FIGmassTransport

\noindent
Figure 1: \unifracLegend

\noindent
Figure 2: \dirtpilesLegend

\clearpage
